---
title: "PM566 Final Project"
author: "Liying Deng"
embed-resources: true
---

# Description {.tabset}

For my public health analytics project, I leverage data-driven insights to explore tobacco use patterns in the United States. Tobacco use remains a significant public health concern, with prevalence and usage patterns varying widely across states due to factors like demographics, cultural norms, public health initiatives, and socioeconomic conditions. Using data from Alaska and Alabama, this project highlights how tobacco behaviors differ in two demographically and geographically distinct regions of the U.S.

Through this study, I investigate key questions about tobacco use trends, focusing on both traditional products such as cigarettes and smokeless tobacco, as well as newer alternatives like e-cigarettes. The analysis covers annual usage trends from 2017 to 2019, demographic differences such as age, gender, and education, and comparisons between the two states. While the study does not evaluate specific smoking cessation strategies, it provides critical insights that can guide the design of future public health initiatives and evidence-based smoking cessation campaigns.

Through in-depth analysis and data visualization, we will discover:

1.  The differences in Cigarette use patterns between a geographically isolated state (Alaska) and a historically tobacco-favoring southeastern state (Alabama).
2.  Trends in the prevalence of specific tobacco products, including e-cigarettes, across different years and demographic groups.
3.  Insights into how age, gender, and education levels influence tobacco usage rates, offering opportunities for targeted interventions.
4.  A comparison of smoking trends and patterns between Alaska and Alabama, highlighting regional disparities in tobacco consumption.

This project not only examines the unique tobacco use behaviors in these two states but also demonstrates the value of data in guiding the development of tailored public health strategies. By identifying usage patterns and trends, this analysis serves as a foundation for future initiatives aimed at reducing tobacco consumption and improving health outcomes. Dive deeper into the findings to explore how data shapes effective public health solutions.

# Download the Written Report {.tabset}

# Interactive visualizations {.tabset}

```{r,echo=FALSE,message=FALSE}
library(dplyr)
library(data.table) 
library(tidyverse)
library(ggplot2)
library(knitr)
library(patchwork)
Tob.Use<- data.table::fread("/Users/kristydeng/Downloads/wsas-xwh5.csv")
Tob.Use <- Tob.Use %>%
  rename(
    state = locationabbr,
    Location = locationdesc
  )
Tob.Use <- Tob.Use %>%
  filter(year %in% c(2017, 2018, 2019))
Tob.Use_Filter <- Tob.Use %>%
  filter(!is.na(data_value))
```

### The trends of Cigarette Use between Alabama and Alaska:

The data visualization reveals distinct trends in adult cigarette use in Alaska and Alabama between 2017 and 2019. In 2017, Alaska exhibited a significantly lower percentage of cigarette users compared to Alabama. However, by 2018, both states showed comparable usage rates, with a notable increase in Alaska and slight stabilization in Alabama. This trend continued into 2019, where cigarette use rates remained closely aligned between the two states.

```{r,echo=FALSE}
filtered_data <- Tob.Use_Filter %>%
  filter(state %in% c("AL", "AK"), year >= 2017, year <= 2019) %>%
  filter(topicdesc %in% c("Cigarette Use (Adults)"))
ggplot(filtered_data, aes(x = as.factor(year), y = data_value, fill = state)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ topicdesc, scales = "free_y") +
  labs(
    title = "Cigarette Use trend in Alabama and Alaska (2017-2019)",
    x = "Year",
    y = "Percentage of Users",
    fill = "State"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1)) 
```

### The Tobacco Use by Education Level in Alabama and Alaska

The tobacco use rate differs significantly depending on the education level. Generally, individuals with education levels below the 12th grade show higher average tobacco usage compared to those with higher education levels.In Alaska, individuals with below 12th-grade education have an average tobacco usage of 27.11%, whereas in Alabama, it is 19.66%. The higher usage in Alaska suggests that lower educational attainment is more strongly associated with tobacco use in Alaska than in Alabama.The boxplot shows that tobacco use rates for individuals with below 12th-grade education have greater variability in Alaska compared to Alabama, indicating wider disparities in usage.The mean tobacco usage among individuals with exactly 12th-grade education is similar in both states, with Alaska having a mean of 15.84% and Alabama having a mean of 14.72%. The boxplot shows comparable ranges for both states, suggesting that individuals who completed high school in both states exhibit similar tobacco use behavior.Individuals with education levels greater than 12th grade have the lowest average tobacco use in both states, with Alaska at 6.94% and Alabama at 8.81%. The relatively low values and minimal variation suggest that higher education correlates with reduced tobacco use across both states. Overall, The boxplot highlight that education level plays a significant role in tobacco consumption. Lower education levels are associated with higher tobacco usage rates, while those with higher education levels (greater than 12th grade) exhibit the lowest tobacco use.Alaska tends to have a slightly higher average usage compared to Alabama across different education levels, particularly among those with lower educational attainment. Public Health Implications: These findings suggest that education level is a key factor in tobacco use behavior. Public health efforts in Alaska and Alabama should focus on providing targeted tobacco cessation programs for individuals with lower education levels, especially those who did not complete high school. Additionally, the data supports the need for educational campaigns to emphasize the health risks of tobacco use, particularly targeting populations with lower education levels.

```{r,echo=FALSE}
edu_summary <- Tob.Use_Filter %>%
  filter(year >= 2017, year <= 2019, 
         state %in% c("AL", "AK"), 
         education != "All Grades") %>%
  group_by(state, education) %>%
  summarise(
    mean_usage = mean(data_value, na.rm = TRUE),
    sd_usage = sd(data_value, na.rm = TRUE),
    count = n(),
    .groups = "drop"
  )
ggplot(Tob.Use_Filter %>% filter(year >= 2017, year <= 2019, 
                          state %in% c("AL", "AK"), 
                          education != "All Grades"),  # Exclude "All Grades"
       aes(x = education, y = data_value, fill = state)) +
  geom_boxplot() +
  labs(
    title = "Tobacco Use Based on Education Levels in Alabama and Alaska (2017-2019)",
    x = "Education Level",
    y = "User Percentage",
    fill = "State"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### The tobacco use vary between males and females in Alabama and Alaska from 2018 to 2019.

In both Alabama and Alaska, the percentages of male and female tobacco users are relatively similar for traditional cigarettes. This indicates that smoking prevalence is not heavily skewed towards a specific gender for this product. However, for newer alternatives such as e-cigarettes, slight gender differences are observed, with males typically showing higher usage rates than females. This aligns with broader trends of males being early adopters of newer tobacco technologies. From 2018 to 2019, the percentages of male and female users remain relatively stable, suggesting that gender-based tobacco use patterns in both states have not significantly shifted within this timeframe. This stability indicates that external factors such as public health campaigns or new regulations may not have significantly impacted gendered usage rates during this period. Smoking cessation programs should account for these gender dynamics. For instance, campaigns targeting e-cigarette cessation might need to emphasize male-focused outreach strategies, as they constitute a larger proportion of users.

```{r,echo=FALSE}
summary_by_gender <- Tob.Use_Filter %>%
  filter(gender != "Overall") %>%
  group_by(state, gender, topicdesc) %>%
  summarise(
    mean_usage = mean(data_value, na.rm = TRUE),
    sd_usage = sd(data_value, na.rm = TRUE),
    count = n(),
    .groups = "drop"
  )
filtered_data_gender <- Tob.Use_Filter %>%
  filter(state %in% c("AL", "AK"), year >= 2018, year <= 2019) %>%
  filter(topicdesc %in% c("Smokeless Tobacco Use (Adults)", 
                          "E-Cigarette Use (Adults)", 
                          "Cigarette Use (Adults)")) %>%
  filter(gender %in% c("Female", "Male"))  # Exclude "overall"
ggplot(filtered_data_gender, aes(x = as.factor(year), y = data_value, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge",width = 0.7) +
  facet_grid(topicdesc ~ state, scales = "free_y") +  
  labs(
    title = "Tobacco Use by Gender in Alabama and Alaska (2018-2019)",
    x = "Year",
    y = "Percentage of Users",
    fill = "Gender"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
    strip.text.y = element_text(size = 4, margin = margin(20, 20, 20, 20)) 
  )
```


### The mean of different age in different type of tobacco use in Alabama and Alaska 

Cigarette Use Dominates Across All Age Groups. Cigarette use remains the most prevalent form of tobacco consumption in both Alabama and Alaska, with the highest usage observed among individuals aged 25 to 44 years. This pattern is consistent across the three years analyzed.For the younger age group (18 to 24 years), cigarette use is still prominent, suggesting the need for targeted interventions to prevent early smoking initiation. E-Cigarette Usage Increasing Among Younger Adults.E-cigarette usage is higher among younger adults (18 to 24 years) compared to older age groups, reflecting its appeal as a newer alternative to traditional tobacco. This trend is particularly significant given the growing concerns about e-cigarette addiction among youth.Public health campaigns should prioritize anti-smoking and e-cigarette prevention programs targeting younger adults and teenagers. Educational initiatives highlighting the health risks of e-cigarettes are essential to curb their growing popularity. The government can strengthen tobacco taxation and enforce restrictions on tobacco advertising and availability to discourage use.

```{r,echo=FALSE}
plot_tobacco_by_year <- function(data, year) {
  filtered_data <- data %>%
    filter(state %in% c("AL", "AK"), 
           year == !!year, 
           age != "All Ages")  # Exclude "All Ages"
  summary_data <- filtered_data %>%
    group_by(state, topicdesc, age) %>%
    summarise(
      mean_usage = mean(data_value, na.rm = TRUE),
      count = n(),
      .groups = "drop"
    )
  ggplot(summary_data, aes(x = age, y = mean_usage, fill = topicdesc)) +
    geom_bar(stat = "identity", position = "dodge",) +
    labs(
      title = paste("Mean Tobacco Use (", year, ")", sep = ""),
      x = "Age Group", 
      y = "Mean Tobacco Use (%)",
      fill = "Tobacco Type"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 5, hjust = 0.5, size = 4,face = "bold"),
           axis.title.y = element_text(size = 5.5, face = "bold"))
}

plot_2017 <- plot_tobacco_by_year(Tob.Use_Filter, 2017)
plot_2018 <- plot_tobacco_by_year(Tob.Use_Filter, 2018)
plot_2019 <- plot_tobacco_by_year(Tob.Use_Filter, 2019)

combined_plot <- plot_2017 / plot_2018 / plot_2019
print(combined_plot)
```

